---
title: "Human_Breast_Cancer"
author: "Jonathan"
date: "2024-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

library setup
```{r}
library(SpatialExperiment)
library(spdep)
library(scran)
library(scuttle)
library(dplyr)
library(ggspavis)
library(spdep)
library(scater)
library(pheatmap)
```

data setup
```{r}
dir <- file.path("/Users/jonathanhuang/Desktop/Boston_University/Research_Rotation/ST_Project/MSAB_Research_Rotation/Human_Breast_Cancer/FFPE_Files")
samples <- file.path(dir, "outs")
FFPE<- read10xVisium(samples = samples)
spe_FFPE <- FFPE[, colData(FFPE)$in_tissue == 1]
spe_FFPE_f <- addPerCellQC(spe_FFPE)
```

Overview + Quality Control
```{r}
plotSpots(FFPE)
head(colData(FFPE))
head(rowData(FFPE))
head(spatialCoords(FFPE))

spe_FFPE <- FFPE[, colData(FFPE)$in_tissue == 1]
dim(spe_FFPE)

# calculate per-spot QC metrics and store in colData
spe_FFPE_f <- addPerCellQC(spe_FFPE)
head(colData(spe_FFPE_f))
```


Local Moran's I Implentation for number of UMI
```{r}
umi <- colData(spe_FFPE_f)$sum
id <- rownames(spatialCoords(spe_FFPE_f))
oid <- order(id)

nb_k6 <- knn2nb(knearneigh(spatialCoords(spe_FFPE_f), k=6))
resI_k6 <- localmoran(umi, nb2listw(nb_k6))
nb_k18 <- knn2nb(knearneigh(spatialCoords(spe_FFPE_f), k=18))
resI_k18 <- localmoran(umi, nb2listw(nb_k18))
nb_k36 <- knn2nb(knearneigh(spatialCoords(spe_FFPE_f), k=36))
resI_k36 <- localmoran(umi, nb2listw(nb_k36))

lmi_data_k6 <- printCoefmat(data.frame(resI_k6[oid,], row.names=rownames(spatialCoords(spe_FFPE_f))[oid]),
 check.names=FALSE)
lmi_data_k18 <- printCoefmat(data.frame(resI_k18[oid,], row.names=rownames(spatialCoords(spe_FFPE_f))[oid]),
 check.names=FALSE)
lmi_data_k36 <- printCoefmat(data.frame(resI_k36[oid,], row.names=rownames(spatialCoords(spe_FFPE_f))[oid]),
 check.names=FALSE)

colData(spe_FFPE_f)$Ii_sum_k6 <- lmi_data_k6$Ii
colData(spe_FFPE_f)$Ii_sum_k18 <- lmi_data_k18$Ii
colData(spe_FFPE_f)$Ii_sum_k36 <- lmi_data_k36$Ii

colData(spe_FFPE_f)$Ii_sum_k6 <- ifelse(colData(spe_FFPE_f)$Ii_sum_k6 < 0,0, colData(spe_FFPE_f)$Ii_sum_k6)
colData(spe_FFPE_f)$Ii_sum_k18 <- ifelse(colData(spe_FFPE_f)$Ii_sum_k18 < 0,0, colData(spe_FFPE_f)$Ii_sum_k18)
colData(spe_FFPE_f)$Ii_sum_k36 <- ifelse(colData(spe_FFPE_f)$Ii_sum_k36 < 0,0, colData(spe_FFPE_f)$Ii_sum_k36)

plotSpots(spe_FFPE_f, annotate = "Ii_sum_k6", pal = c("grey", "navy"),point_size = 1.2)+
  labs(title = "Sum of UMI with High AutoCorr (K = 6)")
plotSpots(spe_FFPE_f, annotate = "Ii_sum_k18", pal = c("grey", "navy"),point_size = 1.2)+
  labs(title = "Sum of UMI with High AutoCorr (K = 18)")
plotSpots(spe_FFPE_f, annotate = "Ii_sum_k36", pal = c("grey", "navy"),point_size = 1.2)+
  labs(title = "Sum of UMI with High AutoCorr (K = 36)")
```


Local Moran's I Implentation for Detected Features
```{r}
umi <- colData(spe_FFPE_f)$detected
id <- rownames(spatialCoords(spe_FFPE_f))
oid <- order(id)

nb_k6 <- knn2nb(knearneigh(spatialCoords(spe_FFPE_f), k=6))
resI_k6 <- localmoran(umi, nb2listw(nb_k6))
nb_k18 <- knn2nb(knearneigh(spatialCoords(spe_FFPE_f), k=18))
resI_k18 <- localmoran(umi, nb2listw(nb_k18))
nb_k36 <- knn2nb(knearneigh(spatialCoords(spe_FFPE_f), k=36))
resI_k36 <- localmoran(umi, nb2listw(nb_k36))

lmi_data_k6 <- printCoefmat(data.frame(resI_k6[oid,], row.names=rownames(spatialCoords(spe_FFPE_f))[oid]),
 check.names=FALSE)
lmi_data_k18 <- printCoefmat(data.frame(resI_k18[oid,], row.names=rownames(spatialCoords(spe_FFPE_f))[oid]),
 check.names=FALSE)
lmi_data_k36 <- printCoefmat(data.frame(resI_k36[oid,], row.names=rownames(spatialCoords(spe_FFPE_f))[oid]),
 check.names=FALSE)

colData(spe_FFPE_f)$Ii_detected_k6 <- lmi_data_k6$Ii
colData(spe_FFPE_f)$Ii_detected_k18 <- lmi_data_k18$Ii
colData(spe_FFPE_f)$Ii_detected_k36 <- lmi_data_k36$Ii

colData(spe_FFPE_f)$Ii_detected_k6 <- ifelse(colData(spe_FFPE_f)$Ii_detected_k6 < 0,0, colData(spe_FFPE_f)$Ii_detected_k6)
colData(spe_FFPE_f)$Ii_detected_k18 <- ifelse(colData(spe_FFPE_f)$Ii_detected_k18 < 0,0, colData(spe_FFPE_f)$Ii_detected_k18)
colData(spe_FFPE_f)$Ii_detected_k36 <- ifelse(colData(spe_FFPE_f)$Ii_detected_k36 < 0,0, colData(spe_FFPE_f)$Ii_detected_k36)

plotSpots(spe_FFPE_f, annotate = "Ii_detected_k6", pal = c("grey", "navy"),point_size = 1.2)+
   labs(title = "Number of Detected Features with High AutoCorr (K = 6)")
plotSpots(spe_FFPE_f, annotate = "Ii_detected_k18", pal = c("grey", "navy"),point_size = 1.2)+
   labs(title = "Number of Detected Features with High AutoCorr (K = 18)")
plotSpots(spe_FFPE_f, annotate = "Ii_detected_k36", pal = c("grey", "navy"),point_size = 1.2)+
   labs(title = "Number of Detected Features with High AutoCorr (K = 36)")
```


non-spatial clustering
```{r}
spe_2 <- logNormCounts(spe_FFPE_f)

dec <- modelGeneVar(spe_2)
top_hvgs <- getTopHVGs(dec, prop = 0.1)

# compute PCA
set.seed(123)
spe_2 <- runPCA(spe_2, subset_row = top_hvgs)
# compute UMAP on top 50 PCs
set.seed(123)
spe_2 <- runUMAP(spe_2, dimred = "PCA")
# update column names
colnames(reducedDim(spe_2, "UMAP")) <- paste0("UMAP", 1:2)


# graph-based clustering
set.seed(123)
k <- 42
g <- buildSNNGraph(spe_2, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
table(clus)

# store cluster labels in column 'label' in colData
colLabels(spe_2) <- factor(clus)

# plot clusters in spatial x-y coordinates
plotSpots(spe_2, annotate = "label", 
          pal = "libd_layer_colors",point_size = 1.2)

# plot clusters in PCA reduced dimensions
plotDimRed(spe_2, plot_type = "PCA", 
           annotate = "label", pal = "libd_layer_colors")

# plot clusters in UMAP reduced dimensions
plotDimRed(spe_2, plot_type = "UMAP", 
           annotate = "label", pal = "libd_layer_colors")

#Local Moran's I through Sum of UMI


# umi <- colData(spe_2)$sum
# id <- rownames(spatialCoords(spe_2))
# nb <- knn2nb(knearneigh(spatialCoords(spe_2), k=18))
# oid <- order(id)
# resI <- localmoran(umi, nb2listw(nb))
# 
# lmi_data <- printCoefmat(data.frame(resI[oid,], row.names=rownames(spatialCoords(spe_2))[oid]),
#  check.names=FALSE)
# colData(spe_2)$Ii_sum <- lmi_data$Ii
# plotSpots(spe_2, annotate = "Ii_sum",
#           pal = c("grey", "navy"), point_size = 1.2) + 
# labs(title = "K = 18")



# set gene names as row names for easier plotting
rownames(spe_2) <- rowData(spe_2)$symbol

# test for marker genes
markers <- findMarkers(spe_2, test = "binom", direction = "up")
# returns a list with one DataFrame per cluster
markers

# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
best_set
logFCs <- getMarkerEffects(best_set)
logFCs
pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))
```


Filtering for high autocorrlation (Ii > 0.5)
```{r}
spe_FFPE_filter_0 <- spe_2[,colData(spe_2)$Ii_sum >0.5]
plotSpots(spe_FFPE_filter_0, annotate = "label", 
          pal = "libd_layer_colors",point_size = 1) + 
labs(title = "Sum of UMI with High AutoCorr (Li > 0.5 with K = 18)")

#Analzying the proportion of high Auto Corr with the clusters
total_corr <- length(colData(spe_FFPE_filter_0)$Ii_sum)

hc_c1 <- sum(colData(spe_FFPE_filter_0)$label == 1)
p_c1 <- hc_c1/total_corr*100

hc_c2 <- sum(colData(spe_FFPE_filter_0)$label == 2)
p_c2 <- hc_c2/total_corr*100

hc_c3 <- sum(colData(spe_FFPE_filter_0)$label == 3)
p_c3 <- hc_c3/total_corr*100

hc_c4 <- sum(colData(spe_FFPE_filter_0)$label == 4)
p_c4 <- hc_c4/total_corr*100

hc_c5 <- sum(colData(spe_FFPE_filter_0)$label == 5)
p_c5 <- hc_c5/total_corr*100

hc_c6 <- sum(colData(spe_FFPE_filter_0)$label == 6)
p_c6 <- hc_c6/total_corr*100

hist(as.numeric(colData(spe_FFPE_filter_0)$label), xlab = "Cluster Type", main = paste("Cluster Distribution (Li > 0.5)"), right = FALSE, breaks = c(1,2,3,4,5,6))
```


Filtering for high autocorrlation (Ii > 1.5)
```{r}
spe_FFPE_filter_1 <- spe_2[,colData(spe_2)$Ii_sum >1.5]
plotSpots(spe_FFPE_filter_1, annotate = "label", 
          pal = "libd_layer_colors", point_size = 1) + 
labs(title = "Sum of UMI with High AutoCorr (Li > 1.5 with K = 18)")

#Analzying the proportion of high Auto Corr with the clusters
total_corr <- length(colData(spe_FFPE_filter_1)$Ii_sum)

hc_c1 <- sum(colData(spe_FFPE_filter_1)$label == 1)
p_c1 <- hc_c1/total_corr*100

hc_c2 <- sum(colData(spe_FFPE_filter_1)$label == 2)
p_c2 <- hc_c2/total_corr*100

hc_c3 <- sum(colData(spe_FFPE_filter_1)$label == 3)
p_c3 <- hc_c3/total_corr*100

hc_c4 <- sum(colData(spe_FFPE_filter_1)$label == 4)
p_c4 <- hc_c4/total_corr*100

hc_c5 <- sum(colData(spe_FFPE_filter_1)$label == 5)
p_c5 <- hc_c5/total_corr*100

hc_c6 <- sum(colData(spe_FFPE_filter_1)$label == 6)
p_c6 <- hc_c6/total_corr*100

hist(as.numeric(colData(spe_FFPE_filter_1)$label), xlab = "Cluster Type", main = paste("Cluster Distribution (Li > 1.5)"), right = FALSE, breaks = c(1,2,3,4,5,6))
```

Filtering for high autocorrlation (Ii > 2.5)
```{r}
spe_FFPE_filter <- spe_2[,colData(spe_2)$Ii_sum >2.5]
plotSpots(spe_FFPE_filter,  annotate = "label", 
          pal = "libd_layer_colors", point_size = 1) + 
labs(title = "Sum of UMI with High AutoCorr (Li > 2.5 with K = 18)")

#Analzying the proportion of high Auto Corr with the clusters
total_corr <- length(colData(spe_FFPE_filter)$Ii_sum)

hc_c1 <- sum(colData(spe_FFPE_filter)$label == 1)
p_c1 <- hc_c1/total_corr*100

hc_c2 <- sum(colData(spe_FFPE_filter)$label == 2)
p_c2 <- hc_c2/total_corr*100

hc_c3 <- sum(colData(spe_FFPE_filter)$label == 3)
p_c3 <- hc_c3/total_corr*100

hc_c4 <- sum(colData(spe_FFPE_filter)$label == 4)
p_c4 <- hc_c4/total_corr*100

hc_c5 <- sum(colData(spe_FFPE_filter)$label == 5)
p_c5 <- hc_c5/total_corr*100

hc_c6 <- sum(colData(spe_FFPE_filter)$label == 6)
p_c6 <- hc_c6/total_corr*100

hist(as.numeric(colData(spe_FFPE_filter)$label), xlab = "Cluster Type", main = paste("Cluster Distribution (Li > 2.5)"), right = FALSE, breaks = c(1,2,3,4,5,6))
```

Filtering for high autocorrlation (Ii > 3.5)
```{r}
spe_FFPE_filter_3 <- spe_2[,colData(spe_2)$Ii_sum >3.5]
plotSpots(spe_FFPE_filter_3,  annotate = "label", 
          pal = "libd_layer_colors", point_size = 1) + 
labs(title = "Sum of UMI with High AutoCorr (Li > 3.5 with K = 18)")

#Analzying the proportion of high Auto Corr with the clusters
total_corr <- length(colData(spe_FFPE_filter_3)$Ii_sum)

hc_c1 <- sum(colData(spe_FFPE_filter_3)$label == 1)
p_c1 <- hc_c1/total_corr*100

hc_c2 <- sum(colData(spe_FFPE_filter_3)$label == 2)
p_c2 <- hc_c2/total_corr*100

hc_c3 <- sum(colData(spe_FFPE_filter_3)$label == 3)
p_c3 <- hc_c3/total_corr*100

hc_c4 <- sum(colData(spe_FFPE_filter_3)$label == 4)
p_c4 <- hc_c4/total_corr*100

hc_c5 <- sum(colData(spe_FFPE_filter_3)$label == 5)
p_c5 <- hc_c5/total_corr*100

hc_c6 <- sum(colData(spe_FFPE_filter_3)$label == 6)
p_c6 <- hc_c6/total_corr*100

hist(as.numeric(colData(spe_FFPE_filter_3)$label), xlab = "Cluster Type", main = paste("Cluster Distribution (Li > 3.5)"), right = FALSE, breaks = c(1,2,3,4,5,6))
```

Filtering for high autocorrlation (Ii > 4.5)
```{r}
spe_FFPE_filter <- spe_2[,colData(spe_2)$Ii_sum >4.5]
plotSpots(spe_FFPE_filter,  annotate = "label", 
          pal = "libd_layer_colors", point_size = 1) + 
labs(title = "Sum of UMI with High AutoCorr (Li > 4.5 with K = 18)")

#Analzying the proportion of high Auto Corr with the clusters
total_corr <- length(colData(spe_FFPE_filter)$Ii_sum)

hc_c1 <- sum(colData(spe_FFPE_filter)$label == 1)
p_c1 <- hc_c1/total_corr*100

hc_c2 <- sum(colData(spe_FFPE_filter)$label == 2)
p_c2 <- hc_c2/total_corr*100

hc_c3 <- sum(colData(spe_FFPE_filter)$label == 3)
p_c3 <- hc_c3/total_corr*100

hc_c4 <- sum(colData(spe_FFPE_filter)$label == 4)
p_c4 <- hc_c4/total_corr*100

hc_c5 <- sum(colData(spe_FFPE_filter)$label == 5)
p_c5 <- hc_c5/total_corr*100

hc_c6 <- sum(colData(spe_FFPE_filter)$label == 6)
p_c6 <- hc_c6/total_corr*100

hist(as.numeric(colData(spe_FFPE_filter)$label), xlab = "Cluster Type", main = paste("Cluster Distribution (Li > 4.5)"), right = FALSE, breaks = c(1,2,3,4,5,6))
```


Differential expression testing (without high-autocorrelation)
```{r}

# set gene names as row names for easier plotting
rownames(spe_2) <- rowData(spe_2)$symbol

# test for marker genes
markers <- findMarkers(spe_2, test = "binom", direction = "up")
# returns a list with one DataFrame per cluster
markers

# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)

pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))

# plot log-transformed normalized expression of top genes for one cluster
top_genes <- head(rownames(interesting))

plotExpression(spe_2, x = "label", features = top_genes)
```


Differential expression testing (with high-autocorrelation Ii > 0.5)
```{r}
# set gene names as row names for easier plotting
rownames(spe_FFPE_filter_0) <- rowData(spe_FFPE_filter_0)$symbol

# test for marker genes
markers <- findMarkers(spe_FFPE_filter_0, test = "binom", direction = "up")
# returns a list with one DataFrame per cluster
markers

# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)

pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))

# plot log-transformed normalized expression of top genes for one cluster
top_genes <- head(rownames(interesting))

plotExpression(spe_2, x = "label", features = top_genes)
```


Differential expression testing (with high-autocorrelation Ii > 1.5)
```{r}
# set gene names as row names for easier plotting
rownames(spe_FFPE_filter_1) <- rowData(spe_FFPE_filter_1)$symbol

# test for marker genes
markers <- findMarkers(spe_FFPE_filter_1, test = "binom", direction = "up")
# returns a list with one DataFrame per cluster
markers

# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)

pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))

# plot log-transformed normalized expression of top genes for one cluster
top_genes <- head(rownames(interesting))

plotExpression(spe_2, x = "label", features = top_genes)
```

Differential Expression on Region of interest
```{r}
x <- as.data.frame(spatialCoords(spe_FFPE_filter_1))
row <- 11000 #smaller than col
col <- 12000 #greater than row
test1 <- x[x$pxl_col_in_fullres < col & x$pxl_row_in_fullres < row,]

try1 <- spe_FFPE_filter_1[,which(row.names(colData(spe_FFPE_filter_1)) %in% row.names(test1))]

plotSpots(try1, annotate = "label", 
          pal = "libd_layer_colors", point_size = 10)


spe_2 <- logNormCounts(spe_FFPE_f)

dec <- modelGeneVar(spe_2)
top_hvgs <- getTopHVGs(dec, prop = 0.1)

# compute PCA
set.seed(123)
spe_2 <- runPCA(spe_2, subset_row = top_hvgs)
# compute UMAP on top 50 PCs
set.seed(123)
spe_2 <- runUMAP(spe_2, dimred = "PCA")
# update column names
colnames(reducedDim(spe_2, "UMAP")) <- paste0("UMAP", 1:2)

spe_2_test <- spe_2
colLabels(spe_2_test) <- ifelse(row.names(colData(spe_2)) %in% row.names(test1),1,2)
colLabels(spe_2_test)

# plot clusters in spatial x-y coordinates
plotSpots(spe_2_test, annotate = "label", 
          pal = "libd_layer_colors",point_size = 1.2) +
labs(title = "Cluster of Interest vs. Others")


# plot clusters in PCA reduced dimensions
plotDimRed(spe_2_test, plot_type = "PCA", 
           annotate = "label", pal = "libd_layer_colors")

# plot clusters in UMAP reduced dimensions
plotDimRed(spe_2_test, plot_type = "UMAP", 
           annotate = "label", pal = "libd_layer_colors")


rownames(spe_2_test) <- rowData(spe_2)$symbol

markers_test <- findMarkers(spe_2_test, test = "binom", direction = "up")

interesting_test <- markers_test[[1]]
interest_test_upper <- interesting_test[order(-interesting_test$logFC.2), ]
sit_top <- as.data.frame(head(interest_test_upper,20))
top_genes <- head(rownames(interest_test_upper),20)

interest_test_bottom <- interesting_test[order(interesting_test$logFC.2), ]
sit_bottom <- as.data.frame(head(interest_test_bottom,20))
bot_genes <- head(rownames(interest_test_bottom),20)
```